Model to study co-evolution of trade an culture

### 2015-04-05 article.tex
correction on introduction and state of the art on models. Left discussion and ccl

%XRC this is wrong because blablabla
As it has been stated before, trades influence cultural transmission by modifying both the goods present in a region and the people consuming these goods. On the one side, analysing separately trade and cultural transmission allows to formalize their inner mechanisms and understand their respective dynamics. On the other hand, in order to analyse the adoption of goods within a province one must take into account both trading strategy and cultural transmission. For example, the modification of good's consumption after the rupture of a connection in the trade system is better explained by knowing the local adoption of this good (cultural transmission) and the prices of this good (trade).

%XRC we propose a model of coevolution to explore the deep links between trade and cultural transmission in the ancient Mediterranean
We propose in this article a framework to study the interaction between cultural transmission and trade. This framework should be able first to reproduce well known dynamics of both cultural transmission and trading. Second, to be useful, this framework should be able to compare results obtained by taking cultural or trade into consideration. In order to build this framework we view trade events as part of the culture. The price of a good are then viewed as a cultural trait. This traits can be transmitted and modified along different rules. By this mean, traditional works previously expressed in a trading framework, can be expressed in a cultural evolution framework. Therefore, the dynamics observed on various cultural transmission mechanisms can be straightforwardly compared to dynamics obtained by trading mechanisms. In order to demonstrate the feasibility of this approach we propose a minimal implementation able to reproduce both cultural transmission dynamics and trade dynamics. Additionally we will analyse the results obtained with tools from both economy and cultural evolution.

\jm{to integrate}
Ultimately, we aim to create a model to study the relation between cultural change and trade networks within the Roman Empire. Examples of questions that could be asked to our model are: What is the relation between distances in the trade network and cultural changes ? Did anti-conformism promote the adaptation of new products coming from remote provinces ? In order to answer these questions the model used has to take into account both cultural effects and trade influence. More precisely, the variations in these two aspects should be modelled and should influence each other.  Therefore, in order to answer the questions targeted we need a co-evolutionary model able to take into account both trade and cultural changes.

@@

The model presented in this work aims to offer a framework for the co-evolution of culture and trades which relies on a low number of parameters. It is not within the scope of this article to answer the questions asked at the beginning of this section as they require specific implementations of the model proposed. However, the article will focus on the presentation of the framework, its justification and more importantly its validation by reproducing previous results obtained on cultural change and trade mechanisms.

@@

\subsection{Models of Trade}

@@

Previous historical and archaeological studies to model trade in the ancient Mediterranean societies have focus on descriptive approaches ~\cite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. 
These works have conclude that the Roman economy was a type of market economy where a variety of small tightly connected markets had loose connections between them~\cite{temin_market_2001,temin_economy_2006,wilson_approaches_2009}. 
Based on these conclusions, further works have attempt to define more exactly which type market economy was in use, i.e. was it close to our market economy or remote ?

\xrc{close or remote in terms of what?}

However, a limitation of the previous works is the impossibility to test the conclusions reached by the authors. The models proposed reflects only the view of the authors on the data analysed. A new type of model would be necessary to go beyond the interpretation, and allow the verification of the conclusions reached. These models should be able to generate data which is then qualitatively compared to the real historical data.

\xrc{what data is being used?}

By this mean, if a factor is really important in the economical network, its removal from the model will lead to data qualitatively different from the historical data. The Agent Based Models (ABM) methodology offers a way to produce this kind of model~\cite{lake_trends_2014}.

\xrc{I would add a paragraph briefly explaining what is an ABM}

Based on this tool a relatively complex and exact modelisation of the prehispanic Pueblo societies has been proposed~\cite{kohler_modelling_2012}. This work offers an ABM integrating numerous effects such as food consumption, production and landscape. The questions studied are  numerous and some works are performed on the economical system and its ties with social transmissions~\cite{kobti_emergence_2006,cockburn_simulating_2013}. However the work done in this project consider a pre-market society in a specific village which makes it difficult to reach more general conclusions, notably applicable to the Roman Empire.

\xrc{The Pueblo model is an example, but there are tons of other ones...why is this chosen? After the paragraph explaining ABM I would say "As an example of its potential, a detailed model of the prehispanic blablabla..."}

\xrc{Didn't Brughmans have a model of roman trade?}

\jm{to integrate}
Among the various works done with this objective, few are studying the interactions between economy and cultural change in the Roman empire~\cite{terpstra_trade_2011,kessler_organization_2007,scheidel_model_2007}. An analysis based on historical documents has revealed the importance of economical and social institutions for the efficiency of trades of grains in the early Roman Empire~\cite{kessler_organization_2007}. The same type of analysis has also revealed the importance of a number of key factors (such as commercial development, population growth and geographical mobility) for the growth of income and population in the Roman Italy~\cite{scheidel_model_2007}. Finally, a micro-economic model has determined the importance of reputation and threat of expulsion from a trade network as mechanisms to maintain the overall efficiency of the trade network~\cite{terpstra_trade_2011}. These works therefore reveal the strong relation between trade and cultural change in the Roman empire.
\xrc{I would move this discussion to the coevolution section? Besides, I would ask something such as "a large percentage of datasets used for this purpose are material objects found in archaeological contexts (i.e. pottery). Most of them are usually represented following a neutral model, but how this affects its weight inside the market system?"}

@@

The model proposed is a Multi-Agent Based model based on agents and goods. The interactions between agents and goods are modelled thanks to the four processes (production, cultural transmission, innovation and trade) which will be explained in further details later. The flexibility of the models comes from the possibility to implement these processes differently depending on the hypothesis tested. However, since all experiments are performed with the same four processes, the effect of the various implementations of one process can be studied easily. Moreover, the operating of the model revolves around the value that agents associate to products. Depending on the question studied, the value can reflect the interest of the agent for the good or it can be the price at which the agent evaluate this good. Since this vector is common to all experiments performed with this model, it is extremely useful for the comparisons of results obtained with different implementation of the key processes.

>>>>

%XRC this is wrong because blablabla
As it has been stated before, trades influence cultural transmission by modifying both the goods present in a region and the people consuming these goods. Certainly, analysing separately trade and cultural transmission allows to formalize their inner mechanisms and understand their respective dynamics. However, in order to analyse the adoption of goods within a province one must take into account both trading strategy and cultural transmission. For example, the modification of good's consumption after the rupture of a connection in the trade system is better explained by knowing the local adoption of this good (cultural transmission) and the prices of this good (trade).

%XRC we propose a model of coevolution to explore the deep links between trade and cultural transmission in the ancient Mediterranean
Ultimately, we aim to create a model to study the relation between cultural change and trade networks within the Roman Empire. Examples of questions that could be asked to our model are: What is the relation between distances in the trade network and cultural changes ? Did anti-conformism promote the adaptation of new products coming from remote provinces ? In order to answer these questions the model used has to take into account both cultural effects and trade influence. More precisely, the variations in these two aspects should be modelled and should influence each other.  Therefore, in order to answer the questions targeted we need a co-evolutionary model able to take into account both trade and cultural changes.

@@

The model presented in this work aims to offer a framework for the co-evolution of culture and trades which relies on a low number of parameters. Moreover, to be useful, the framework is developed with the aim to ease the comparison of results obtained when taking cultural or trade into consideration.  This article will focus on the presentation of the framework, its justification and more importantly its validation by reproducing previous results obtained on cultural change, and the ability to produce trade dynamics. The results obtained when studying trade and cultural change are compared with each other.

@@

\subsection{Models including Trade}

@@

Previous historical and archaeological studies to model trade in the ancient Mediterranean societies have focus on descriptive approaches ~\cite{hopkins_taxes_1980,temin_market_2001,terpstra_trade_2011,temin_economy_2006,wilson_approaches_2009,scheidel_model_2007,kessler_organization_2007}. 
These works have conclude that the Roman economy was a type of market economy where a variety of small tightly connected markets had loose connections between them~\cite{temin_market_2001,temin_economy_2006}. 
Based on these conclusions, further works have attempt to define more exactly the characteristics of this market such, e.g. level of intervention of the state and its impact, or the relation-ship between local and long-distance trade~\cite{wilson_approaches_2009}.

Few works have attempt to model the interactions between economy and cultural change in the Roman empire~\cite{terpstra_trade_2011,kessler_organization_2007,scheidel_model_2007}. An analysis based on historical documents has revealed the importance of economical and social institutions for the efficiency of trades of grains in the early Roman Empire~\cite{kessler_organization_2007}. The same type of analysis has also revealed the importance of a number of key factors (such as commercial development, population growth and geographical mobility) for the growth of income and population in the Roman Italy~\cite{scheidel_model_2007}. Finally, a micro-economic model has determined the importance of reputation and threat of expulsion from a trade network as mechanisms to maintain the overall efficiency of the trade network~\cite{terpstra_trade_2011}. These works therefore reveal the strong relation between trade and cultural change in the Roman empire. However a large percentage of the datasets used for these studies are based on material objects found in archaeological contexts such as pottery. Most of these objects are usually represented following a neutral model, but the weight of this model within a market system is not studied.

An additional limitation of the previous works is the impossibility to test the conclusions reached by the authors. The models proposed reflects only the view of the authors on the historical and archaeological data analysed. A new type of model would be necessary to go beyond the interpretation, and allow the verification of the conclusions reached. These models should be able to generate data relative to trade which are then qualitatively compared to results of analysis of archaeological data such as amphorae or other container used to carry goods. By this mean, if a factor is really important in the economical network, its removal from the model will lead to data qualitatively different from the historical data.

\todo{check that paragraph}
The Agent Based Models (ABM) methodology offers a way to produce this kind of model within a computer~\cite{lake_trends_2014}. An ABM is composed of a group of agents performing actions within a virtual environment. The environment and the behaviour of the agents, despite being an simpler than the real world, are designed to be analogous to their real-world counterpart. The experimenter hand designs the agents, the environment and their interaction, and observes the outcome of all interactions through time in order to compare it to the archaelogical data obtained from the real world. The goal of the experimenter is therefore to find a plausible set of assumptions used during the creation of the ABM produce results close to what is observed in reality.

A first example of its potential is given through a model of the trade network of tableware in the roman east~\cite{brughmans_roman_2014}. Among the parameters tested, preliminary results of this model show that number of links between production site has a strong influence on the results obtained. A second more complete example proposes a detailed model of the prehispanic Pueblo societies~\cite{kohler_modelling_2012}. This work offers an ABM integrating numerous effects such as food consumption, production and landscape. The questions studied are  numerous and some works are performed on the economical system and its ties with social transmissions~\cite{kobti_emergence_2006,cockburn_simulating_2013}. However the work done in this project consider a pre-market society in a specific village which makes it difficult to reach more general conclusions, notably applicable to the Roman Empire.

@@

The model proposed is an ABM based on agents and goods. The interactions between agents and goods are modelled thanks to the four processes (production, cultural transmission, innovation and trade) which will be explained in further details later. The flexibility of the models comes from the possibility to implement these processes differently depending on the hypothesis tested. However, since all experiments are performed with the same four processes, the effect of the various implementations of one process can be studied easily. Moreover, the operating of the model revolves around the value that agents associate to products. Depending on the question studied, the value can reflect the interest of the agent for the good or it can be the price at which the agent evaluate this good. Since this vector is common to all experiments performed with this model, it is extremely useful for the comparisons of results obtained with different implementation of the key processes.
